Think deep for reef resilience

### 2025-09-09 paper/paper_draft.qmd
Add map plot files to paper. Still need to add to discussion
Also regenerate all figures

To assess the impact of reef connectivity on modelled outcomes we summarise inter-reef
connections using weighted connectivity metrics (incoming and outgoing) and the ratio of
source-to-sink connectivity for each reef. Weighted incoming connectivity is calculated by
summing the product of each connection strength incoming to a target reef and the mean
timeseries absolute coral cover (km^2^) of each matching source reef. This metric provides
an approximation of overall larval flow into a reef, with high values indicating the target
reef is supplied high volumes of larvae via strong connections and low values indicating the
target reef has weak (or no) incoming connections or is supplied by small source
populations. Weighted outgoing connectivity is calculated as the product of outgoing
connectivity strength from a target reef and the mean timeseries absolute coral cover
(km^2^) of the target reef, representing the larval flow leaving the target reef population.
A limitation of these weighted connectivity metrics is that they do not incorporate the
proportion of the source reef's population that is reproducing and producing larvae, instead
taking the total coral cover area as an indicator of larval productivity. The ratio of
source-to-sink connectivity ($O:I$) for a target reef represents the strength of the reef's
outgoing larval flow relative to it's incoming larval flow. This is calculated as:
$O:I = \frac{S_O/N_O}{S_I/N_I}$. Where S represents the sum of connectivity pobabilities, N
represents the number of connections, O denotes outgoing connectivity from the target reef
and I denotes incoming connectivity to the target reef. If a reef has no incoming
connections, the $O:I$ equals the $S_O/N_O$ value. Incoming connections in this calculation,
do not directly represent the larval supply gained as the breeding populations from which
connections are coming from are not represented. A source-to-sink ratio greater than 1 (a
value greater than 0 on a log scale) indicates a reef has greater outgoing than incoming
larval flow.

@@

Reefs in each time series cluster were compared to identify characteristics that differ
between Low, Medium and High clusters. These characteristics include reef median depth, mean
time series DHW, ratio of source-to-sink larval connectivity, weighted incoming
connectivity, weighted outgoing connectivity and reef carrying capacity (coral habitable
area).

@@

Across six of the 29 bioregions reefs in the high cover cluster have lower mean DHW, Log-10
carrying capacity and Log-10 weighted outgoing connectivity values than reefs in other
clusters (@sfig-access-esm1-5-bioregion-dhwmean;
@sfig-access-esm1-5-bioregion-carrying-capacity;
@sfig-access-esm1-5-bioregion-weightedoutgoing). There were no consistent patterns across
bioregions in their reefs' Log-10 coral carrying capacity area and cluster assignment
(@sfig-access-esm1-5-bioregion-carrying-capacity). Across most bioregions there were not
consistent differences between clusters in connectivity metrics such as Log-10
source-to-sink connectivity ratio (@sfig-access-esm1-5-bioregion-logsosi) and Log-10
weighted incoming connectivity (@sfig-access-esm1-5-bioregion-weightedincoming).

@@

A Partial Dependence Analysis was conducted using a Random Forest classifier to identify
reef characteristics that influence its temporal behaviour, and thus its cluster assignment
into Low, Medium, or High. Bioregion classification was excluded from this analysis as a
predictor due to its categorical nature providing limited interpretable information, and is
further justified by its low feature importance score of 9%, the lowest of all predictors
(@fig-ts-cluster-pdp; Panel A). Reef median depth and mean DHW level have the highest
feature importance scores, accounting for 25% and 16% predictive importance respectively.
Weighted incoming and outgoing connectivity (Log10) features account for \~13.6% feature
importance. The carrying capacity of a reef accounts for 12.3% feature importance. Reef
source-to-sink ratio (Log10) both had feature importance scores of \~10%.

@@

Analyses for source-to-sink ratio and weighted incoming and outgoing connectivity metrics
suggest that reefs with greater outgoing connectivity and lower incoming connectivity levels
have higher probability of being clustered into the Low cluster category compared to other
reefs. Additionally, two-way analyses revealed an interaction between reef depth and
incoming connectivity in determining cluster category assignment, particularly at the high
depth range (@fig-ts-cluster-pdp-twoway-incoming). Two-way analyses revealed that reefs at
greater than 10m depth are unlikely to be clustered into the Low cluster category,
regardless of their incoming connectivity levels. However, at greater than 10m depth reefs
with higher incoming connectivity are more likely to be assigned to a High cluster, while
reefs with low incoming connectivity are more likely to be assigned to a Medium cover
cluster (@fig-ts-cluster-pdp-twoway-incoming). Two-way analyses, for depth and weighted
outgoing connectivity, revealed a slight interaction with the increased low-cluster
probability for reefs of high outgoing connectivity primarily concentrated in shallow reefs,
while deep reefs with high outgoing connectivity have the greatest probability of being
clustered into a high cover cluster (@sfig-ts-cluster-pdp-twoway-outgoing).

@@

Current reef restoration and intervention investigation efforts on the GBR, such as
deployment of coral seeding and fragment devices, target reefs with suitable sites in the
1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are no explicit
justifications for this depth range, intervention depth is likely constrained by snorkelling
and diving deployment and monitoring logistical accessibility, visual identification of
sites from the surface, in combination with tourism return value. Our results suggest that
of the assessed reefs occurring within the 1-10m depth range (1736 reefs), 57% (991) of
reefs fall into the low and medium clusters (across all GCMs at the GBRMPA management region
scale), with many unable to maintain positive carbonate budgets under future conditions.
Partial dependence analysis revealed a range of depths from 10-15m where the probability of
a reef being clustered into the High cover cluster increased substantially. Additionally,
two-way partial dependence analysis revealed that high incoming connectivity is important
for High cluster assignment for reefs at depths of \>10m. Considering this 10-15m depth
range, and targetting well connected reefs, when assessing possible intervention sites may
reduce the exposure of planted and seeded corals to high thermal stress, while still
maintaining effective logistical access to sites. Managers are also likely to consider the
positive potential for flow on effects due to outgoing connectivity when selecting
intervention sites [@hock2017]. While reefs with high outgoing connectivity levels were
found to have greater probability of being assigned to low cover clusters when assessed
individually (@fig-ts-cluster-pdp-twoway-incoming), within the context of reef depth, reefs
with high depths and high outgoing connectivity levels had the greatest probability of being
clustered into high cover clusters (@sfig-ts-cluster-pdp-twoway-outgoing). These results
suggests that interventions targetting deep reefs will also benefit from high incoming and
outgoing connectivity levels, allowing for high potential for site resilience and outgoing
flow of benefits and recovery. Transient, and difficult to measure, properties such as reef
connectivity are likely to increase uncertainty in intervention planning to a greater degree
than environmental characteristics such as reef depth. Overall, reef intervention planning
is a complex and challenging decision-making process [@anthony2020], and our results suggest
that intervention site depth may be an important factor to additionally consider when
assessing the long-term resilience of sites.

@@

@@

![Weighted outgoing connectivity (in base-10 log scale) for reefs within each bioregion,
separated into clusters assigned by time series clustering. ‘Low’ clusters (green) are
clusters with the lowest median area-relative cover between 2030 and 2060 within each
bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative
cover between 2030 and 2060 within each bioregion. ‘High’ clusters (blue) are clusters with
the highest median area-relative cover between 2030 and 2060 within each bioregion. Solid
lines within each violin represent the median base-10 log of weighted outgoing connectivity
for each cluster and bands represent the distribution of connectivity values within each
cluster. Displays results for ACCESS-ESM1-5
GCM.](../figs/ACCESS-ESM1-5/bioregion/weighted_outgoing_conn_bioregion_violin.png){#sfig-access-esm1-5-bioregion-weightedoutgoing
fig-align="center" width="14.82cm"}

>>>>

To assess the impact of reef connectivity on modelled outcomes we summarise incoming
connectivity (weighted incoming connectivity), outgoing connectivity (outgoing connection strength), the ratio of
source-to-sink connectivity, and larval retention probability for each reef. Weighted incoming connectivity is calculated by
summing the product of each connection strength incoming to a target reef and the mean
timeseries absolute coral cover (km^2^) of each matching source reef. This metric provides
an approximation of overall larval flow into a reef, with high values indicating the target
reef is supplied high volumes of larvae via strong connections and low values indicating the
target reef has weak (or no) incoming connections or is supplied by small source
populations. A limitation of weighted incoming connectivity is it does not incorporate the
proportion of the source reef's population that is reproducing and producing larvae, instead
taking the total coral cover area as an indicator of larval productivity. Outgoing connection strength is calculated as the sum of all outgoing transition probabilities from a target reef. Outgoing connection strength has a theoretical maximum of one (all larvae reach settlement on sink reefs), however the large size of the GBR combined with larval mortality considerations result in outgoing connection strength values much lower than one. The ratio of
source-to-sink connectivity ($O:I$) for a target reef represents the strength of the reef's
outgoing larval flow relative to it's incoming larval flow. This is calculated as:
$O:I = \frac{S_O/N_O}{S_I/N_I}$. Where S represents the sum of connectivity pobabilities, N
represents the number of connections, O denotes outgoing connectivity from the target reef
and I denotes incoming connectivity to the target reef. If a reef has no incoming
connections, the $O:I$ equals the $S_O/N_O$ value. Incoming connections in this calculation,
do not directly represent the larval supply gained as the breeding populations from which
connections are coming from are not represented. A source-to-sink ratio greater than 1 (a
value greater than 0 on a log scale) indicates a reef has greater outgoing than incoming
larval flow. Reef larval retention probability indicates the proportion of spawned larvae from a target reef that remain to settlement.

@@

Reefs in each time series cluster were compared to identify characteristics that differ
between Low, Medium and High clusters. These characteristics include reef median depth, mean
time series DHW, weighted incoming
connectivity, outgoing connection strength, ratio of source-to-sink larval connectivity, larval retention probability and reef carrying capacity (coral habitable
area).

@@

Across bioregions there were no consistent patterns between reef clusters in mean DHW values (@sfig-access-esm1-5-bioregion-dhwmean), Log-10
weighted incoming connectivity (@sfig-access-esm1-5-bioregion-weightedincoming), Log-10 outgoing connection strength (@sfig-access-esm1-5-bioregion-outoingconn), Log-10
source-to-sink connectivity ratio (@sfig-access-esm1-5-bioregion-logsosi) and Log-10 larval retention probability (@sfig-access-esm1-5-bioregion-selfconn). Across six of the 29 bioregions reefs in the high cover cluster have lower Log-10
carrying capacity values than reefs in other
clusters (@sfig-access-esm1-5-bioregion-carrying-capacity).

@@

A Partial Dependence Analysis was conducted using a Random Forest classifier to identify
reef characteristics that influence its temporal behaviour, and thus its cluster assignment
into Low, Medium, or High. Bioregion classification was excluded from this analysis as a
predictor due to its categorical nature providing limited interpretable information, and is
further justified by its low feature importance score of 9%, the lowest of all predictors
(@fig-ts-cluster-pdp; Panel A). Reef median depth, mean DHW level and weighted incoming connectivity (Log10) have the highest
feature importance scores, accounting for 22%, 17% and 14% predictive importance respectively.
The carrying capacity of a reef accounts for 12% feature importance, while outgoing connection strength, source-to-sink ratio and larval retention account for 10%, 9% and 7% respectively.

@@

Analyses for source-to-sink ratio and incoming and outgoing connectivity metrics
suggest that reefs with greater outgoing connectivity and lower incoming connectivity levels
have higher probability of being clustered into the Low cluster category compared to other
reefs. Additionally, two-way analyses revealed an interaction between reef depth and
incoming connectivity in determining cluster category assignment, particularly at the high
depth range (@fig-ts-cluster-pdp-twoway-incoming). Two-way analyses revealed that reefs at
greater than 10m depth are unlikely to be clustered into the Low cluster category,
regardless of their incoming connectivity levels. However, at greater than 10m depth reefs
with higher incoming connectivity are more likely to be assigned to a High cluster, while
reefs with low incoming connectivity are more likely to be assigned to a Medium cover
cluster (@fig-ts-cluster-pdp-twoway-incoming). Two-way analyses, for depth and outgoing connection strength, revealed a slight interaction with deep reefs with high outgoing connectivity having the greatest probability of being
clustered into a high cover cluster (@sfig-ts-cluster-pdp-twoway-outgoing).

@@

Current reef restoration and intervention investigation efforts on the GBR, such as
deployment of coral seeding and fragment devices, target reefs with suitable sites in the
1-10m depth range [@howlett2021; @randall2021; @whitman2024]. While there are no explicit
justifications for this depth range, intervention depth is likely constrained by snorkelling
and diving deployment and monitoring logistical accessibility, visual identification of
sites from the surface, in combination with tourism return value. Our results suggest that
of the assessed reefs occurring within the 1-10m depth range (1736 reefs), 57% (991) of
reefs fall into the low and medium clusters (across all GCMs at the GBRMPA management region
scale), with many unable to maintain positive carbonate budgets under future conditions.
Partial dependence analysis revealed a range of depths from 10-15m where the probability of
a reef being clustered into the High cover cluster increased substantially. Additionally,
two-way partial dependence analysis revealed that high incoming connectivity is important
for High cluster assignment for reefs at depths of \>10m. Considering this 10-15m depth
range, and targetting well connected reefs, when assessing possible intervention sites may
reduce the exposure of planted and seeded corals to high thermal stress, while still
maintaining effective logistical access to sites. Managers are also likely to consider the
positive potential for flow on effects due to outgoing connectivity when selecting
intervention sites [@hock2017]. Within the context of reef depth, reefs
with high depths and high outgoing connectivity levels had the greatest probability of being
clustered into high cover clusters (@sfig-ts-cluster-pdp-twoway-outgoing). These results
suggests that interventions targetting deep reefs will also benefit from high incoming and
outgoing connectivity levels, allowing for high potential for site resilience and outgoing
flow of benefits and recovery. Transient, and difficult to measure, properties such as reef
connectivity are likely to increase uncertainty in intervention planning to a greater degree
than environmental characteristics such as reef depth. Overall, reef intervention planning
is a complex and challenging decision-making process [@anthony2020], and our results suggest
that intervention site depth may be an important factor to additionally consider when
assessing the long-term resilience of sites.

@@

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by median reef depth.
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2007]. Reef centroids are
marked individually for visual clarity. Lighter colours represent greater reef depth values than darker colours. All 2429 reefs included in this
investigation are represented in this figure. Annotations represent major coastal
settlements along the GBR.](../figs/depth_med_input_map.png){#sfig-depth-map
fig-align="center" width="14.82cm"}

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by reef carrying capacity (coral habitable area).
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2007]. Reef centroids are
marked individually for visual clarity. Lighter colours represent greater carrying capacity area values than darker colours. All 2429 reefs included in this
investigation are represented in this figure. Annotations represent major coastal
settlements along the GBR.](../figs/abs_k_area_input_map.png){#sfig-abs-k-area-map
fig-align="center" width="14.82cm"}

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by outgoing connection strength (Log10).
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2007]. Reef centroids are
marked individually for visual clarity. Lighter colours represent higher outgoing connection strength values than darker colours. All 2429 reefs included in this
investigation are represented in this figure. Annotations represent major coastal
settlements along the GBR.](../figs/log_out_strength_input_map.png){#sfig-outgoing-conn-map
fig-align="center" width="14.82cm"}

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by source-to-sink connectivity ratio (Log10).
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2007]. Reef centroids are
marked individually for visual clarity. Lighter colours represent higher source-to-sink values than darker colours. Source-to-sink ratio values above zero (one on normal scale) indicate a reef has greater outgoing than incoming connectivity. All 2429 reefs included in this
investigation are represented in this figure. Annotations represent major coastal
settlements along the GBR.](../figs/log_so_to_si_input_map.png){#sfig-sotosi-conn-map
fig-align="center" width="14.82cm"}

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by larval retention probability (Log10).
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2007]. Reef centroids are
marked individually for visual clarity. Lighter colours represent higher probability of retaining spawned larvae than darker colours. All 2429 reefs included in this
investigation are represented in this figure. Annotations represent major coastal
settlements along the GBR.](../figs/log_self_strength_input_map.png){#sfig-self-conn-map
fig-align="center" width="14.82cm"}

@@

![Outgoing connection strength (in base-10 log scale) for reefs within each bioregion,
separated into clusters assigned by time series clustering. ‘Low’ clusters (green) are
clusters with the lowest median area-relative cover between 2030 and 2060 within each
bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative
cover between 2030 and 2060 within each bioregion. ‘High’ clusters (blue) are clusters with
the highest median area-relative cover between 2030 and 2060 within each bioregion. Solid
lines within each violin represent the median base-10 log of outgoing connection strength
for each cluster and bands represent the distribution of connectivity values within each
cluster. Displays results for ACCESS-ESM1-5
GCM.](../figs/ACCESS-ESM1-5/bioregion/outgoing_strength_conn_bioregion_violin.png){#sfig-access-esm1-5-bioregion-outgoingconn
fig-align="center" width="14.82cm"}

![Larval retention probability (in base-10 log scale) for reefs within each bioregion,
separated into clusters assigned by time series clustering. ‘Low’ clusters (green) are
clusters with the lowest median area-relative cover between 2030 and 2060 within each
bioregion. ‘Medium’ clusters (orange) are clusters with intermediate median area-relative
cover between 2030 and 2060 within each bioregion. ‘High’ clusters (blue) are clusters with
the highest median area-relative cover between 2030 and 2060 within each bioregion. Solid
lines within each violin represent the median base-10 log of larval retention probability
for each cluster and bands represent the distribution of probability values within each
cluster. Displays results for ACCESS-ESM1-5
GCM.](../figs/ACCESS-ESM1-5/bioregion/self_conn_bioregion_violin.png){#sfig-access-esm1-5-bioregion-selfconn
fig-align="center" width="14.82cm"}

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by median number of years exceeding positive carbonate budget threshold.
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2007]. Reef centroids are
marked individually for visual clarity. Reefs that have a median positive carbonate budget year value of 0 years are coloured in grey. All other reefs are coloured according to their median years across threshold values, with lighter colours representing greater number of years with positive carbonate budgets than darker colours. All 2429 reefs included in this
investigation are represented in this figure. This map depicts values from the ACCESS-ESM1-5 GCM. Annotations represent major coastal
settlements along the GBR.](../figs/ACCESS-ESM1-5/median_carbonate_budget_map.png){#sfig-carb-budget-map
fig-align="center" width="14.82cm"}

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by mean time series DHW value.
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2007]. Reef centroids are
marked individually for visual clarity. Lighter colours represent higher mean DHW values than darker colours. All 2429 reefs included in this
investigation are represented in this figure. This map depicts values from the ACCESS-ESM1-5 GCM. Annotations represent major coastal
settlements along the GBR.](../figs/ACCESS-ESM1-5/mean_dhw_input_map.png){#sfig-mean-dhw-map
fig-align="center" width="14.82cm"}

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all
considered reefs (points), coloured by weighted incoming connectivity levels (Log10).
Colour-coded areas outline GBR marine park management regions, where orange represents the
Far Northern management region, green represents Cairns/Cooktown, red covers
Townsville/Whitsunday and blue covers Mackay/Capricorn [@GBRMPA2007]. Reef centroids are
marked individually for visual clarity. Lighter colours represent higher weighted incoming connectivity values than darker colours. All 2429 reefs included in this
investigation are represented in this figure. This map depicts values from the ACCESS-ESM1-5 GCM. Annotations represent major coastal
settlements along the GBR.](../figs/ACCESS-ESM1-5/weighted_incoming_conn_input_map.png){#sfig-incoming-conn-map
fig-align="center" width="14.82cm"}
